Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics

### 2025-07-08 paper/kobakian-cook-anzjs.qmd
typo in equation fixed

type2_colors <- c(
  "Choropleth" = "#fcae91",
  "Hexagon tile" = "#a50f15")

@@

The Australian Cancer Atlas [@atlas] is an online interactive web tool created to explore the burden of cancer on Australian communities. There are many cancer types to be explored individually or aggregated. The Australian Cancer Atlas allows users to explore the patterns in the distributions of cancer statistics over the geographic space of Australia. It uses a choropleth map display and diverging color scheme to draw attention to relationships between neighboring areas. The hexagon tile map may be a useful alternative display to enhance the atlas.

The experiment was conducted using the lineup protocol, a visual inference procedure [@GIIV], to objectively test the effectiveness of the two displays.

@@

<!-- foundation, tradition -->  
Spatial visualisations communicate the distribution of statistics over geographic landscapes. The choropleth map [@EI, @BCM] is a traditional display. It is used to present statistics that have been aggregated on geographic units. Creating a choropleth map involves drawing polygons representing the administrative boundaries, and filling with colour mapped to the value of the statistic. The choropleth map places the statistic in the context of the spatial domain, so that the reader can see whether there are spatial trends, clusters or anomalies. This is important for digesting disease patterns. If there is a trend it may imply that the disease is spreading from one location to another. If there is a cluster, or an anomaly, there may be a localized outbreak of the disease. Aggregating the statistic on administrative units, provides a level of privacy to individuals, while allowing the impact of the disease on the community to be analyzed.

@@

<!-- Cartogram -->
A cartogram is a general solution for adequately displaying a population-based statistic. It transforms the geographic map base to reflect the population in the geographic region, while preserving some aspects of the geographic location. There are several cartogram algorithms [@ACTUC, @CBATCC]; each involves shifting the boundaries of geographic units, using the value of the statistic to increase or decrease the area taken by the geographic unit on the map. The changes to the boundaries result in cartograms that accurately communicate population by map area for each of the geographic units but can result in losing the familiar geographic information. For Australia, the transformations warp the country so that it is no longer recognizable.

<!-- Other cartograms -->
Alternative algorithms make various trade offs between familiar shapes and representation of geographic units. The non-contiguous cartogram method [@NAC] keeps the shapes of geographic units intact, and changes the size of the shape. This method disconnects areas creating empty space on the display losing the continuity of the spatial display of the statistic. The Dorling cartogram [@ACTUC] represents each unit as a circle, sized according to the value of the statistic. The neighbour relationships are mostly maintained by how the circles touch. A similar approach was pioneered by @RSCW, using rectangles that tile to align borders of neighbours [@CDWCS]. There have been thorough reviews of the array of methods, as suitable for cancer atlas displays [@review, @BCM].

@@

<!-- Hexagon tile map -->
The hexagon tile map algorithm, automatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag [@sugarbag].  @fig-thyroid b shows the hexagon tile map, where the map is coloured from substantially below average (blue) to substantially above average (red) rates. The inner city areas have expanded, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is of interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

@@

In order to assess the effectiveness of the hexagon tile map, the lineup protocol [@BCHLLSW09, @GIIV] from visual inference procedures is employed. The approach mirrors classical statistical inference. The procedures for doing a power comparison of competing plot designed, outlined in @GTPCCD, are followed.

@@

Australia is used for the study, with Statistical Area 3 (SA3) [@abs2016] as the geographic units. The results should apply broadly to any other geographic area of interest.

@@

  - Map type: *Choropleth, Hexagon tile*
  - Trend: *South-East to North-West; Locations in three population centres; Locations in multiple population centres, *

@@

```{r}
#| label: fig-exp-design
#| results: "asis" 
#| fig-width: 8 
#| fig-height: 8
#| out-width: 80%
#| fig-cap: "The experimental design used in the study. Participants are allocated to with group A or B, to evaluate either the choropleth or hexagon tile map lineup of each simulated data set. Place in the lineup is recorded as 'Location'."
ggdraw(xlim = c(0,1), ylim = c(0,1)) + draw_plot(rasterGrob(png::readPNG("figures/experiment_design.png")))
```

@@

For each of the 144 individual maps, the values for each geographic area were rescaled to create a similar color scale from deep blue to dark red within each map.
This meant at least one geographic unit was coloured dark blue, and at least one was red, in every map display of every lineup.

For the geographic NW-SE distribution, this resulted in the smallest values of the trend model (blue) occurring in Western Australia, the North West of Australia, and the largest values of the trend model (red) occurring in the South East. This resulted in Tasmania being colored completely red.

@@

The first step in the data cleaning process involved checking that survey responses collected for each participants were only included once in the data set. 
The data cleaning process also involved filtering out participants' who did not provide at least three unique choices when considering each of the twelve lineups. These participants achieved a detection rate of 0. If participants had made various plot choices for the 12 displays they saw they were still included in the dataset.

<!-- Plots will be made in R [@RCore], with the `ggplot2` package [@ggplot2].-->   
### Modelling

$$
\text{logit}(p_{ijk}) = \mu_i + \tau_j + \delta_k + (\tau \delta)_{jk} ~~~~ j=1,2; ~~ k=1,2,3
$$

where $y_{ijk} = 0, 1$ represents whether subject $i$ detected the data plot (1) or did not (0), $\mu_$ is the subject-specific random intercept, $\tau_j$ is the map type effect, $\delta_k$ is the trend model effect. The interaction between map type and trend model allows for any map type effect to differ between trend models. As each participant provides results from 12 lineups, this model can account for each individual participants' abilities with the subject-specific random intercept.

@@

 - Plot choice: the number of the plot that they deemed to be most different from the others.
 - Reason: one of "Clusters of colour", "Colour trend across the areas", "Big differences between neighbouring areas", "All areas have similar colours" or "None of these reasons". 
 - Certainty: how certain that their choice is different from the others, on a scale of 1-5.

@@

@@

@fig-detect-compare displays the average detection rates for the two types of plot separately for each trend model. Each trend model was tested using four repetitions, evaluations on the same data set were seen as either choropleths or hexagon tile maps by each group as specified in @fig-exp-design; the detection rates for each display are connected by a line segment. The Three Cities and All Cities trend models shown in the hexagon tile map allowed viewers to detect the data plot substantially more often than the choropleth counterparts. One replicate for the All Cities group had similar detection rates for both map types, the rate of detection using the choropleth map was much higher than other replicates. Surprisingly, participants could also detect the gradual spatial trend in the NW-SE group from the hexagon tile map. We expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs slightly better, or equally as well.

@@

```{r}
#| label: tbl-desc-stats
#| results: "asis"
types <- c("Choro.", "", "Hex.", "")

d %>% group_by(type, trend) %>%
  summarise(m = as.character(round(mean(detect), 2)),
      std.dev = as.character(round(sd(detect), 2))) %>% 
  mutate(std.dev = ifelse(std.dev == 0.5, "(0.50)", paste0("(", std.dev, ")"))) %>%   mutate(m = ifelse(m == 0.4, "0.40", m)) %>% 
  gather(stat, value, m, std.dev) %>% 
  pivot_wider(names_from = "trend", values_from = "value") %>% 
  arrange(type) %>% ungroup() %>% 
  mutate(Type = types) %>% select(Type, `NW-SE`, `Three Cities`, `All Cities`) %>%
  knitr::kable(., format = "latex", align = "lccc", booktabs = TRUE, 
    linesep = c("", "\\addlinespace"),
    caption = "The mean and standard deviation of the rate of detection for each trend model, calculated for the choropleth and hexagon tile map displays.")
```

@@

When a choropleth map display is used, the predicted detection rate for the Three Cities trend, `r round(exp(0.07-3.41)/(1+exp(0.07-3.41)),2)`; this is extremely low, especially compared to the NW-SE trend of `r round(exp(0.07)/(1+exp(0.07)),2)`.
When the All Cities trend is presented in a choropleth display the predicted probability of detection is `r round(exp(0.07-1.34)/(1+exp(0.07-1.34)),2)`.
The hexagon tile map has a substantially high detection rate for the display of a Three Cities trend `r round(exp(0.46+0.07-3.41+2.44)/(1+exp(0.46+0.07-3.41+2.44)),2)` and All Cities trend `r round(exp(0.46+0.07-1.34+1.16)/(1+exp(0.46+0.07-1.34+1.16)),2)`.

```{r}
#| label: tbl-detect-glmer1
#| results: "asis"
# Mixed effects models
glmer1 <- glmer(detect ~ type*trend + (1|contributor), 
              family = binomial, data = d)

@@

tidy(glmer1) %>%
  mutate_at(.vars = c("estimate", "std.error"), round, 2) %>% 
  mutate(p.value = round(p.value, digits=2)) %>% 
  rowwise() %>%
  mutate(sig = case_when(
  p.value <= 0.001 ~ "$^{***}$",
  p.value <= 0.01 ~  "$^{**}$",
  p.value <= 0.05 ~  "$^{*}$",
  p.value <= 0.01 ~  "$^{.}$",
  TRUE ~ "$^{ }$")) %>%
  ungroup() %>% 
  filter(!is.na(std.error)) %>%
  mutate(term = glmer_terms) %>% 
  select(Term = term, 
    Est. = estimate, 
    Sig. = sig,
    `Std. Error` = std.error, 
    `P val` = p.value) %>% 
  knitr::kable(format = "latex", escape = FALSE, align= "rrlrr", 
    booktabs = T, linesep = c("", "\\addlinespace", "", "\\addlinespace", ""), 
    caption = "The model output for the generalised linear mixed effect model for detection rate. This model considers the type of display, the trend model hidden in the data plot, and accounts for contributor performance.")
```

@@

@fig-beeswarm shows horizontally jittered dot plots to contrast the time taken by participants to evaluate each lineup when viewing each type of display. The time are also separated by trend model and whether the data plot was detected or not detected. The time taken to complete an evaluation ranged from milliseconds to 60 seconds. The average time taken for type of display is shown as a large colored dot on each plot. when considering the heights of the green and orange dots, there is little difference in the average time taken to read a choropleth or hexagon tile map. Comparing the same colored dot across each trend model row, there is a slight increase in the time taken to correctly detected the data plot in the hexagon tile map lineup, but little difference in evaluation time for the choropleth display. However, there were substantially less correct detections for choropleth lineups for the Three cities and All Cities trends.

```{r}
#| label: fig-beeswarm
#| fig-cap: "The distribution of the time taken (seconds) to submit a response for each combination of trend, whether the data plot was detected, and type of display, shown using horizontally jittered dotplots. The colored point indicates average time taken for each map type. Although some participants take just a few seconds per evaluation, and some take as much as mcuh as 60 seconds, but there is very little difference in time taken between map types."
#| fig-height: 6
#| fig-width: 5
#| out-width: 80%
s <- d %>% group_by(type, trend, detect_f) %>%
  summarise(m=median(time_taken), 
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75))
library(ggbeeswarm)
ggplot() + 
  geom_quasirandom(data=d, aes(x=type, y=time_taken), alpha=0.9) + 
  #geom_hline(data=s, aes(yintercept=m, color=type)) +
  geom_point(data=s, aes(x=type, y=m, color=type), size=5, alpha=0.7) +
  #geom_errorbar(data=s, aes(x=type, ymin=q1, ymax=q3, color=type), width=0.3, size=2) +
  scale_color_manual("", values = type_colors) +
  facet_grid(trend~detect_f) + 
  ylab("Time taken (seconds)") + xlab("") +
  theme(legend.position="none")
```

@@

Participants provided their level of certainty regarding their choice using a five point scale. 
Unlike the accuracy and speed of responses that were derived during the data processing phase, this was a subjective
assessment by the participant prompted by the question: ‘How certain are you about your choice?’.
@fig-certainty shows the amount of times participants provided each level of certainty. This was separated for each combination of trend models and display type, and colored depending on whether a participant correctly detected the data plot in the lineup.
Participants often chose 4 or 5 when viewing the population related trends in the choropelth display, even though they were often incorrect when viewing an All Cities trend and overwhelmingly incorrect for the Three Cities trend. This shows overconfidence in their detection ability when using a choropleth map display. Participants were less likely to be certain when their choice was incorrect and they were viewing a hexagon tile map.
For each trend model, participants were more likely to doubt their choice and choose 1 or 2 in the hexagon tile map displays, even though many had made the correct choice.

@@

# devtools::install_github("haleyjeppson/ggmosaic")
library(ggmosaic)
d |> 
  mutate(Detected = factor(detect_f, 
    levels = c("Detected? Yes", "Detected? No"), 
    labels = c("Detected? Yes", "Detected? No"))) |>
  mutate(type = fct_recode(type, "Choropleth" = "Choro.", "Hexagon tile" = "Hex.")) |>
  ggplot() +
    geom_mosaic(aes(x = certainty, fill = type)) +  
    scale_fill_manual("", values = type2_colors) +
    scale_y_continuous("proportion", breaks=seq(0,1,0.25)) +
    facet_grid(Detected ~ trend) +
    theme(aspect.ratio = 0.8, 
      axis.text = element_text(size="10"),
      legend.position = "bottom")
```

@@

Participants were asked why they had made their plot choice and were able to select from a set of suggested reasons. 
"Color trend across the areas" was the most common selection for NW-SE trend displays (@tbl-reason).

The reasons chosen by participants from the list provided to them varied more when viewing choropleth displays than the hexagon tile map.
The hexagon tile map displays resulted in "Clusters of color" as the most common choice made by participants.

@@

```{r}
#| label: tbl-reason
#| results: "asis"
# Qualitative analysis of reason
d %>% 
  mutate(reason = ifelse(reason =="0.0", "no reason", reason)) %>% 
  mutate(Detected = ifelse(detect_f == "Detected? Yes", "Yes", "No"),
    Trend = trend) %>% 
  group_by(Trend, Detected, type) %>% 
  count(reason) %>% 
  mutate(prop = round(n/sum(n), 2), r_prop = paste0(reason, ":", prop)) %>% 
  top_n(1, n) %>% summarise(reasons = paste(reason, collapse=", ")) %>% 
  pivot_wider(names_from = c("type"), values_from = c("reasons")) %>% 
  ungroup() %>% 
  knitr::kable(., format = "latex", booktabs = TRUE,
    linesep = c("", "\\addlinespace", "", "\\addlinespace",""),
    caption = "The amount of participants that selected each reason for their choice of plot when looking at each trend model shown in Choropleth and Hexagon Tile maps. The facets show whether or not the choice was correct.") %>% 
  #kable_styling(bootstrap_options = "hold_position") %>% 
  collapse_rows(., columns = 1)
```

@@

The color scaling applied in Three cities and All cities displays resulted in the rural areas of the real data plot appearing more blue or yellow than the other plots in the lineups.
Due to the consistent coloring of rural areas in a choropleth display, the choice "All areas have similar colors" was most common reason for a participants choice. The All Cities displays colored the inner-city areas of all capital cities more red, this was observable to participants and explains the equal choice of the city clusters or rural color consistency. 
Choosing "Clusters of colour" was expected when participants viewed the Hexagon tile map display of the All Cities and Three Cities distributions. It was unexpected that it was also the most common reason for the NW-SE hexagon tile map displays. 
Due to the spatial covariance introduced in the smoothing, groups of similarly colored hexagons were present in all of the hexagon tile map displays. All Cities and Three Cities distributions of real data trends had distinctly different patterns or red inner-city areas, while some of the plots in each lineup may have shared similar features.

@@

@@

>>>>

type2_colors <- c(
  "choropleth" = "#fcae91",
  "hexagon tile" = "#a50f15")

@@

The Australian Cancer Atlas [@atlas] is an online interactive web tool created to explore the burden of cancer on Australian communities. There are many cancer types to be explored individually or aggregated. The Australian Cancer Atlas allows users to explore the patterns in the distributions of cancer statistics over the geographic space of Australia. It uses a choropleth map display and diverging color scheme to draw attention to relationships between neighbouring areas. The hexagon tile map may be a useful alternative display to enhance the atlas.

The experiment was conducted using the lineup protocol, a visual inference procedure [@BCHLLSW09;@GIIV], that can be used to objectively test the effectiveness of the two displays [@GTPCCD]. A lineup embeds the data plot among a field of null data plots, and an independent observer is asked to select the most different plot. It was shown by @VVSIALM to be an effective way to conduct a hypothesis test where a plot is treated as a test statistic. If the data plot is selected it is analogous to a rejection of the null hypothesis (no structure), and the observed data plot is unlikely to arise from the null scenario.

@@

<!-- foundation, tradition -->  
Spatial visualisations communicate the distribution of statistics over geographic landscapes. The choropleth map [@EI;@BCM] is a traditional display. It is used to present statistics that have been aggregated on geographic units. Creating a choropleth map involves drawing polygons representing the administrative boundaries, and filling with colour mapped to the value of the statistic. The choropleth map places the statistic in the context of the spatial domain, so that the reader can see whether there are spatial trends, clusters or anomalies. This is important for digesting disease patterns. If there is a linear trend it may imply a relationship between disease and geographic location. If there is a cluster, or an anomaly, there may be a localized outbreak of the disease. <!--Aggregating the statistic on administrative units, provides a level of privacy to individuals, while allowing the impact of the disease on the community to be analyzed.-->

@@

<!-- Cartogram -->
A cartogram is a general solution for adequately displaying a population-based statistic. It transforms the geographic map base to reflect the population in the geographic region, while preserving some aspects of the geographic location. There are several cartogram algorithms [@ACTUC;@CBATCC]; each involves shifting the boundaries of geographic units, using the value of the statistic to increase or decrease the area taken by the geographic unit on the map. The changes to the boundaries result in cartograms that accurately communicate population by map area for each of the geographic units but can result in losing the familiar geographic information. For Australia, the transformations warp the country so that it is no longer recognizable.

<!-- Other cartograms -->
Alternative algorithms make various trade offs between familiar shapes and representation of geographic units. The non-contiguous cartogram method [@NAC] keeps the shapes of geographic units intact, and changes the size of the shape. This method disconnects areas creating empty space on the display losing the continuity of the spatial display of the statistic. The Dorling cartogram [@ACTUC] represents each unit as a circle, sized according to the value of the statistic. The neighbour relationships are mostly maintained by how the circles touch. A similar approach was pioneered by @RSCW, using rectangles that tile to align borders of neighbours [@CDWCS]. There have been thorough reviews of the array of methods, as suitable for cancer atlas displays [@review;@BCM].

@@

<!-- Hexagon tile map -->
The hexagon tile map algorithm, automatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag [@sugarbag].  @fig-thyroid b shows the hexagon tile map, where the map is coloured from substantially below average (blue) to substantially above average (red) rates. The inner city areas have expanded, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is of interest to know how well the spatial distribution patterns are seen from this display, in comparison to how they are seen from the choropleth.

@@

In order to assess the effectiveness of the hexagon tile map, the lineup protocol [@BCHLLSW09;@GIIV] from visual inference procedures is employed. The approach mirrors classical statistical inference. The procedures for doing a power comparison of competing plot designed, outlined in @GTPCCD, are followed.

@@

Australia is used for the study, with Statistical Area 3 (SA3) [@abs2016] as the geographic units. The results should apply broadly to any other geographic areas of interest, if there are large differences in area and population size.

@@

  - Map type: *choropleth, hexagon tile*
  - Trend: *South-East to North-West; Locations in three population centres; Locations in multiple population centres, *

@@

```{r}
#| label: fig-exp-design
#| results: "asis" 
#| fig-width: 6 
#| fig-height: 4
#| out-width: 100%
#| fig-cap: "The experimental design used in the study. Participants are allocated to with group A or B, to evaluate either the choropleth or hexagon tile map lineup of each simulated data set. The text 'loc' refers to the location of the data plot in the lineup."
#ggdraw(xlim = c(0,1), ylim = c(0,1)) + #draw_plot(rasterGrob(png::readPNG("figures/experiment_design.png")))

expt_design <- d |>
  count(group, trend, location, type, replicate)

ggplot(expt_design, ) +
  geom_tile(aes(x=group, y=1, fill=type)) +
  geom_text(aes(x=group, y=1, label = paste("loc:", location))) + 
  facet_grid(replicate~trend, axes = "all", labeller = labeller(replicate = label_both, trend = label_both)) +
  ylab("") +
  scale_fill_manual("", values = type_colors) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

@@

For each of the 144 individual maps, the values for each geographic area were rescaled to create a similar colour scale from deep blue to dark red within each map.
This meant at least one geographic unit was coloured dark blue, and at least one was red, in every map display of every lineup.

For the geographic NW-SE distribution, this resulted in the smallest values of the trend model (blue) occurring in Western Australia, the North West of Australia, and the largest values of the trend model (red) occurring in the South East. This resulted in Tasmania being coloured completely red.

@@

@@

 - **Plot choice**: the number of the plot that they deemed to be most different from the others.
 - **Reason**: one of "Clusters of colour", "Colour trend across the areas", "Big differences between neighbouring areas", "All areas have similar colours" or "None of these reasons". 
 - **Certainty**: how certain that their choice is different from the others, on a scale of 1-5.

@@

Data is checked to ensure that survey responses collected for each participants were only included once. Technically it is possible to submit results more than once if the submit button is clicked multiple times in short sequence. Participants who did not finish the evaluation of all lineups or clicked through without providing their evaluation are filtered.

<!-- Plots will be made in R [@RCore], with the `ggplot2` package [@ggplot2].-->

### Modelling

$$
\text{logit}(p_{ijk}) = \mu_i + \tau_j + \delta_k + (\tau \delta)_{jk}
$$

where $y_{ijk} = 0, 1$ represents whether subject $i$ detected the data plot (1) or did not (0), $\mu_i, ~i=1, \dots n$ is the subject-specific random intercept, $n$ is the number of subjects, $\tau_j, ~j=1,2$ is the map type effect, $\delta_k, ~k=1,2,3$ is the trend model effect. The interaction between map type and trend model allows for any map type effect to differ between trend models. As each participant provides results from 12 lineups, this model can account for each individual participants' abilities with the subject-specific random intercept.

@@

@fig-detect-compare displays the average detection rates for the two types of plot separately for each trend model. Each trend model was tested using four repetitions, evaluations on the same data set were seen as either choropleths or hexagon tile maps by each group as specified in @fig-exp-design; the detection rates for each display are connected by a line segment. The Three Cities and All Cities trend models shown in the hexagon tile map allowed viewers to detect the data plot substantially more often than the choropleth counterparts. One replicate for the All Cities group had similar detection rates for both map types - the rate of detection using the choropleth map was much higher than other replicates. Surprisingly, participants could also detect the gradual spatial trend in the NW-SE group from the hexagon tile map. We expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs slightly better, or equally as well.

@@

```{r}
#| label: tbl-desc-stats
#| tbl-cap: "The mean and standard deviation of the rate of detection for each trend model, calculated for the choropleth and hexagon tile map displays."
types <- c("Choro.", "", "Hex.", "")

d %>% group_by(type, trend) %>%
  summarise(m = as.character(round(mean(detect), 2)),
      std.dev = as.character(round(sd(detect), 2))) %>% 
  mutate(std.dev = ifelse(std.dev == 0.5, "(0.50)", paste0("(", std.dev, ")"))) %>%   mutate(m = ifelse(m == 0.4, "0.40", m)) %>% 
  gather(stat, value, m, std.dev) %>% 
  pivot_wider(names_from = "trend", values_from = "value") %>% 
  arrange(type) %>% ungroup() %>% 
  mutate(Type = types) %>% select(Type, `NW-SE`, `Three Cities`, `All Cities`) %>%
  knitr::kable(., format = "latex", align = "lccc", booktabs = TRUE, 
    linesep = c("", "\\addlinespace"))
```

@@

When a choropleth map display is used, the predicted detection rate for the Three Cities trend, `r round(exp(0.07-3.41)/(1+exp(0.07-3.41)),2)` is extremely low, especially compared to the NW-SE trend of `r round(exp(0.07)/(1+exp(0.07)),2)`.
When the All Cities trend is presented in a choropleth display the predicted probability of detection is `r round(exp(0.07-1.34)/(1+exp(0.07-1.34)),2)`.
The hexagon tile map has a substantially high detection rate for the display of a Three Cities trend `r round(exp(0.46+0.07-3.41+2.44)/(1+exp(0.46+0.07-3.41+2.44)),2)` and All Cities trend `r round(exp(0.46+0.07-1.34+1.16)/(1+exp(0.46+0.07-1.34+1.16)),2)`.

```{r}
#| label: tbl-detect-glmer1
#| tbl-cap: "Parameter estimates of the fitted model fit for  detection rate. All terms are statistically significant  ($^{**}=0.01$, $^{***}=0.001$)."
# Mixed effects models
glmer1 <- glmer(detect ~ type*trend + (1|contributor), 
              family = binomial, data = d)

@@

tidy(glmer1) %>%
  mutate_at(.vars = c("estimate", "std.error"), round, 2) %>% 
  mutate(p.value = round(p.value, digits=2)) %>% 
  rowwise() %>%
  mutate(sig = case_when(
  p.value <= 0.001 ~ "$^{***}$",
  p.value <= 0.01 ~  "$^{**}$",
  p.value <= 0.05 ~  "$^{*}$",
  p.value <= 0.01 ~  "$^{.}$",
  TRUE ~ "$^{ }$")) %>%
  ungroup() %>% 
  filter(!is.na(std.error)) %>%
  mutate(term = glmer_terms) %>% 
  select(Term = term, 
    Est. = estimate,
    `Std. Err.` = std.error, 
    `P-val.` = p.value, 
    Sig. = sig) %>% 
  knitr::kable(format = "latex", escape = FALSE, align= "rrlrr", 
    booktabs = T, linesep = c("", "\\addlinespace", "", "\\addlinespace", ""))
```

@@

@fig-beeswarm shows horizontally jittered dot plots to contrast the time taken by participants to evaluate each lineup when viewing each type of display. The time are also separated by trend model and whether the data plot was detected or not detected. The time taken to complete an evaluation ranged from milliseconds to 60 seconds. The average time taken for type of display is shown as a large coloured dot on each plot. When considering the heights of the green and orange dots, there is little difference in the average time taken to read a choropleth or hexagon tile map. Comparing the same coloured dot across each trend model row, there is a slight increase in the time taken to correctly detected the data plot in the hexagon tile map lineup, but little difference in evaluation time for the choropleth display. However, there were substantially less correct detections for choropleth lineups for the Three cities and All Cities trends.

```{r}
#| label: fig-beeswarm
#| fig-cap: "The distribution of the time taken (seconds) to submit a response for each combination of trend, whether the data plot was detected, and type of display, shown using horizontally jittered dotplots. The coloured point indicates average time taken for each map type. Although some participants take just a few seconds per evaluation, and some take as much as 60 seconds, but there is very little difference in time taken between map types."
#| fig-height: 6
#| fig-width: 5
#| out-width: 80%
s <- d %>% group_by(type, trend, detect_f) %>%
  summarise(m=median(time_taken), 
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75))
library(ggbeeswarm)
ggplot() + 
  geom_quasirandom(data=d, aes(x=type, y=time_taken), alpha=0.9) + 
  #geom_hline(data=s, aes(yintercept=m, color=type)) +
  geom_point(data=s, aes(x=type, y=m, color=type), size=5, alpha=0.7) +
  #geom_errorbar(data=s, aes(x=type, ymin=q1, ymax=q3, color=type), width=0.3, size=2) +
  scale_color_manual("", values = type_colors) +
  facet_grid(trend~detect_f) + 
  ylab("Time taken (seconds)") + xlab("") +
  theme(legend.position="none")
```

@@

Participants provided their level of certainty regarding their choice using a five point scale. Unlike the accuracy and speed of responses that were derived during the data processing phase, this was a subjective assessment by the participant prompted by the question: "How certain are you about your choice?". @fig-certainty shows a mosaic plot summarising how participants reported their certainty with their decision. The plots are made separately for each combination of trend models and whether the data plot was detected or not. Colour indicates the type of map display. The proportion of subjects evaluating the choropleth version increases as certainty increases, which suggests overconfidence in their detection ability when using a choropleth map display. Participants were less likely to be certain when their choice was incorrect and they were viewing a hexagon tile map.

@@

# devtools::install_github("haleyjeppson/ggmosaic")
library(ggmosaic)
d |> 
  mutate(Detected = factor(detect_f, 
    levels = c("Detected? Yes", "Detected? No"), 
    labels = c("Detected? Yes", "Detected? No"))) |>
  mutate(type = fct_recode(type, "choropleth" = "Choro.", "hexagon tile" = "Hex.")) |>
  ggplot() +
    geom_mosaic(aes(x = certainty, fill = type)) +  
    scale_fill_manual("", values = type2_colors) +
    scale_y_continuous("proportion", breaks=seq(0,1,0.25)) +
    facet_grid(Detected ~ trend) +
    theme(aspect.ratio = 0.8, 
      axis.text = element_text(size="10"),
      legend.position = "bottom")
```

@@

Participants were asked why they had made their plot choice and were able to select from a set of suggested reasons. 
"Colour trend across the areas" was the most common selection for NW-SE trend displays (@tbl-reason).

The reasons chosen by participants from the list provided to them varied more when viewing choropleth displays than the hexagon tile map.
The hexagon tile map displays resulted in "Clusters of colour" as the most common choice made by participants.

@@

```{r}
#| label: tbl-reason
#| tbl-cap: "The main reason that participants wrote for their choice of plot when looking at each trend model shown in choropleth and hexagon tile maps."
# Qualitative analysis of reason
d %>% 
  mutate(reason = ifelse(reason =="0.0", "no reason", reason)) %>% 
  mutate(Detected = ifelse(detect_f == "Detected? Yes", "Yes", "No"),
    Trend = trend) %>% 
  group_by(Trend, Detected, type) %>% 
  count(reason) %>% 
  mutate(prop = round(n/sum(n), 2), r_prop = paste0(reason, ":", prop)) %>% 
  top_n(1, n) %>% summarise(reasons = paste(reason, collapse=", ")) %>% 
  pivot_wider(names_from = c("type"), values_from = c("reasons")) %>% 
  ungroup() %>% 
  knitr::kable(., format = "latex", booktabs = TRUE,
    linesep = c("", "\\addlinespace", "", "\\addlinespace","")) %>% 
  #kable_styling(bootstrap_options = "hold_position") %>% 
  collapse_rows(., columns = 1)
```

@@

The colour scaling applied in Three cities and All cities displays resulted in the rural areas of the real data plot appearing more blue or yellow than the other plots in the lineups.
Due to the consistent colouring of rural areas in a choropleth display, the choice "All areas have similar colours" was most common reason for a participants choice. The All Cities displays coloured the inner-city areas of all capital cities more red, this was observable to participants and explains the equal choice of the city clusters or rural colour consistency. 
Choosing "Clusters of colour" was expected when participants viewed the Hexagon tile map display of the All Cities and Three Cities distributions. It was unexpected that it was also the most common reason for the NW-SE hexagon tile map displays. 
Due to the spatial covariance introduced in the smoothing, groups of similarly coloured hexagons were present in all of the hexagon tile map displays. All Cities and Three Cities distributions of real data trends had distinctly different patterns or red inner-city areas, while some of the plots in each lineup may have shared similar features.

@@

# Supplementary materials and reproducibility {-}

@@

# Ethics Declaration {-}
